---
title: 'RNA-Seq Analysis'
author: 'Andrew Powers'
output: 
  html_document:
    code_folding: hide
---

```{r load libraries in}
knitr::opts_chunk$set(echo=T, warning=F, message=F)
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library("DESeq2"))
suppressPackageStartupMessages(library(pheatmap))
library(RColorBrewer)
library(DT)
options(digits=2)

```

# Introduction

Alzheimer's is a debilitating disease that affects about 6 million people living in the United States, 
according to the Alazheim'ers Association. One in three seniors die from some form of dementia. As
our populations continue to live longer and we make more advances in science, the amount of people
affected by dementia will keep increasing.
We have been tasked with running a RNA-Seq analysis upon this publicly available data from Recount2.

## Study Explanation

The data is a publicly available dataset that can be found on the Recount2 website. The project
ID is SRP02656. They collected 12 samples in total, with 6 samples coming form normal to early stage
Alzheimers and the other 6 coming from late stage alzheimer's patients.

```{r load in data}
#recount_brain <- add_metadata()
#interest <- subset(recount_brain)
#url  <- download_study("SRP026562")

# Load in the data from the downloaded file
load(file.path('SRP026562', 'rse_gene.Rdata'))
#convert rse_gene to something small
rse <- rse_gene
# Add the additional labels of Alzheimers and Control to the dataset to test on
condition <- c(rep('control',6),rep('alzheimer',6))
rse@colData$condition <- factor(condition, levels=c('control', 'alzheimer'),
  labels=c('control', 'alzheimer'))
# Add the additional labels for Sex, so we can check to see if it is important
rse@colData$sex <- ifelse(sum(grepl('female', rse@colData$characteristics)), 'female', 'male')
```

```{r get the DSEq loaded in}
# load the data into a DESeq2
ddsSE <- DESeqDataSet(rse, design = ~ condition)
keep_dds <- rowSums(counts(ddsSE)) >= 10
dds <- ddsSE[keep_dds,]

```


# Quality Control

## Heatmap Between/Within Samples


```{r heatmap stuff}

ddsSE <- estimateSizeFactors(ddsSE)
vst.dds <- vst(ddsSE, blind=TRUE)
vst.mat.all <- assay(vst.dds)
vst.cor.all <- cor(vst.mat.all)
metadata <- data.frame(condition, row.names=rownames(vst.cor.all))
metadata <- cbind(metadata, rse@colData$sex)
colnames(metadata)[2] <- 'Sex'
pheatmap(vst.cor.all, annotation = metadata)
```
Explain the heatmap


## PCA Plot
```{r ma plot}
# MA plot for differences
plotPCA(vst.dds, intgroup="condition")+
  labs(title="PCA Plot colored by Condition")+
  theme(plot.title = element_text(hjust=0.5))


plotPCA(vst.dds, intgroup="sex")+
  labs(title="PCA Plot colored by Sex")+
  theme(plot.title = element_text(hjust=0.5))
```


## DE Analysis

```{r test}
dds_run <- DESeq(dds)

plotDispEsts(dds_run, main='Dispersion Plot')

res <- results(dds_run,
               contrast= c("condition", "alzheimer", "control"),
               alpha=0.5,
               lfcThreshold=0.32)

top.five <- res %>%
  data.frame() %>% 
  slice_max(order_by=stat, n=5)


res <- lfcShrink(dds_run,
                 contrast= c("condition", "alzheimer", "control"),
                 res = res,
                 type="ashr")


plotMA(res, ylim=c(-8,8))

```

### Top Significant DEGs

```{r Result exploration}
#sub the normalized counts out
heat.colors <- brewer.pal(6, "YlOrRd")
 sig.genes <- res %>%
  data.frame() %>%
  filter(padj < 0.05)
sig.gene.counts <- vst.mat.all[rownames(sig.genes),]
pheatmap(sig.gene.counts,
         color=heat.colors,
         cluster_rows=T,
         show_rownames=F,
         annotation = metadata,
         scale="row")


deg.top <- sig.genes %>% 
  slice_max(log2FoldChange, n=10)
deg.bottom <- sig.genes %>%
  slice_min(log2FoldChange, n=10)

deg.all <- rbind(deg.top, deg.bottom)
deg.gene.counts <- vst.mat.all[rownames(deg.all),]
pheatmap(deg.gene.counts,
         color=heat.colors,
         cluster_rows=T,
         show_rownames=T,
         annotation = metadata,
         scale="row",
         main="Top 10 Upregulated and Downregulated Genes Heatmap")
```

```{r volcano plot}
res_all  <- res.df %>%
  rownames_to_column(var = "ensgene") %>%
  mutate(threshold = padj < 0.05)

ggplot(res_all, aes(x=log2FoldChange, y=-log10(padj), color=threshold))+
  geom_point()+
  labs(x='Log2 Fold Change', y='-Log10 Adjusted p-value', title='Volcano Plot')+
  theme(legend.position = "none", plot.title = element_text(hjust=0.5))
```



